---
title: "'The Best Interests of the Child'"
subtitle: "A Dangerous Cliche"
description: "(estimated time to read:  4 mins)"
preview: no-cliches.jpg
author:
  - name: Dan Swart
    url: https://uncommonboardadvice.netlify.app
date: 2025-04-07
base_url: https://uncommonboardadvice.netlify.app/
categories:
  - School Governance
  - Board Practice
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

>Oue aim is to solve problems...avoid trouble.

## Beyond "The Best Interests of the Child": Finding Real Solutions

This seemingly innocent phrase can create serious problems for our boards. It's another form of misrepresentation as it assumes we know far more than we possibly can. This expression has migrated from courtrooms into our schools, typically entering education through systems perceived as more rigorous. This simple-sounding phrase creates significant challenges when we use it to guide our decisions.

## The Origin and Misapplication

This legal standard began in family law for custody decisions when children faced specific risks. In that narrow context, with individual children and particular circumstances, courts could at least attempt to evaluate tangible factors. Even there, many legal experts criticize it for being too subjective.

In our education system, the phrase becomes even more problematic when:

1. Applied across thousands of different students simultaneously
2. Not connected to specific, measurable results
3. Used by people without deep knowledge of individual children
4. Used as justification for almost any position

## This Phrase Can Stop Productive Conversation

"In the best interests of the child" functions primarily as a rhetorical device that:

1. **Halts discussion**: Who among us would argue against children's interests?
2. **Claims moral authority**: Makes the speaker appear to be the true child advocate
3. **Avoids specificity**: Never clearly defines what those interests actually include
4. **Resists verification**: Creates a standard that cannot be proven wrong

The phrase effectively serves as a blank check that stops conversation rather than encouraging it.


## The Real-World Problems We Face

For us as board members, this phrase creates several practical challenges:

### 1. It Lacks Clear Definition

What exactly constitutes a child's "best interest"? Does it mean:

- Higher test scores?
- Emotional well-being?
- Future earning potential?
- Current happiness?
- Social skills development?
- Alignment with family values?
- Preparation for citizenship?

Without specifying which interests take priority (and acknowledging they sometimes conflict), the phrase provides no real guidance for decision-making.

### 2. It Ignores Necessary Resource Tradeoffs

Educational decisions always involve tradeoffs. Our resources are limited, and focusing on one area means less attention to others:

```{r tradeoffs-table, echo=FALSE, warning=FALSE, message=FALSE}
library(magrittr)   # for pipe operator

tradeoffs <- data.frame(
  Priority = c("Academic rigor", "Social-emotional support", "Arts education", "Athletics", "Vocational training"),
  `Potential Benefits` = c(
    "Content mastery, college readiness",
    "Emotional skills, relationship skills",
    "Creative expression, cultural appreciation",
    "Physical health, teamwork",
    "Practical skills, career preparation"
  ),
  `Potential Costs` = c(
    "Stress, less time for other activities",
    "May mean less academic content coverage",
    "May mean fewer resources for core subjects",
    "Budget costs, injury risks",
    "May mean fewer resources for traditional academics"
  )
)

knitr::kable(tradeoffs, "html") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%
  kableExtra::column_spec(1, bold = TRUE)

```

The "best interests" phrase implies we can maximize everything simultaneously, which simply isn't possible.

### 3. It Assumes Universal Agreement

The phrase presumes we all share the same understanding of what's best for children, when in reality:

* Families in our communities have different values and priorities

* Our communities have different needs and resources

* Cultures represented in our districts have different expectations

* Individual children have different strengths and challenges

What one parent considers essential, another may view as harmful or unnecessary.

### 4. It Often Conceals Adult Agendas

Perhaps most troubling, the phrase frequently masks adult interests:

- Political agendas extending beyond education

- Professional advancement opportunities

- Avoidance of difficult conversations

- Protection from criticism or legal action

When these adult interests hide behind child-centered language, honest debate becomes impossible.


### The Financial Accountability Problem
From a budget perspective, "in the best interests of the child" is particularly problematic because it:

1. Rejects fiscal constraints as less important than children's needs

2. Creates moral pressure to approve spending without proper scrutiny

3. Makes cost-benefit analysis appear uncaring or 'anti-child'

4. Expands district missions beyond core educational functions

As board members, we have a responsibility to manage public funds that cannot be abandoned for undefined "interests."


## A Better Approach: Specificity and Transparency

Instead of this vague, emotional phrase, we should require:

**Specific goals:** "This program aims to improve reading proficiency by X percentage points"

**Clear measurements:** "Success will be evaluated using these three indicators..."

**Honest tradeoffs:** "Implementing this initiative means we'll need to reduce resources for that..."

**Evidence:** "Research indicates this approach works because..."

**Resource requirements:** "This will require X dollars and Y staff hours..."

## Examples of More Productive Discussions

Instead of accepting:

"We need this new social-emotional curriculum in the best interests of our children."

We should ask:

Is this properly the district's responsibility? 

Is our district the best provider for this need?

What specific results will improve? 

How will we measure success? 

What tradeoffs are involved?

What evidence supports its effectiveness? What resources will it require?"


### When This Phrase Appears in Discussions

When someone uses "the best interests of the child," we as board members should respond with:

"Could you clarify which specific interests you're addressing?"

"How would we measure whether we're advancing those interests?"

"What evidence suggests this approach serves those interests better than alternatives?"

"What tradeoffs should we consider in pursuing these particular interests?"

"How do you balance competing interests that might conflict?"

The phrase "in the best interests of the child" signals either unclear thinking or potential manipulation. While children's needs must remain central to our educational decisions, vague appeals to undefined "interests" hinder rather than help our governance.


## Imagine the Difference...

How much more effective our boards could be if we moved from empty rhetoric to substantive dialogue:

By requiring specificity, evidence, and measurable results

By discussing tradeoffs honestly and openly

By establishing clear priorities that reflect our community's values

Together, we can create decision-making processes that truly serve our students, families, and communities.




### If you liked this post or found it valuable - share it with others.


### Your Thoughts?

### Share your perspective in the comments below. Use the 'Sign in with GitHub' button to verify you're not a robot.


