---
title: "The Education-Government Complex"
description: "(estimated time to read:  4 mins)"
preview: tx-legislature.jpg
author:
  - name: Dan Swart
    url: https://uncommonboardadvice.netlify.app
date: 2025-04-11
base_url: https://uncommonboardadvice.netlify.app/
categories:
  - Education Policy
  - Texas Politics
  - School Governance
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



<div class="callout-important">
<strong>Wow!</strong> Talk about making things more complicated!!  You can avoid a lot of the cost and trouble these 'solutions' bring a district.  The "education-government complex" isn't inevitable - it's a created market that benefits everyone except those it should serve: teachers and students.
</div>




## The Education Money Machine: Following the Profit Trail

Have you noticed something troubling in our education system? It's become a perfect money-making machine — not for our students or teachers, but for companies selling "solutions" to problems created by government mandates.

Here's how this cycle works against our districts:

1. Lawmakers create extremely detailed mandates (like the TEKS in Texas)
2. These mandates are too complex for teachers to track on their own
3. Companies swoop in with expensive "solutions" to help schools comply
4. Our districts must buy these products to meet state requirements
5. More mandates create more markets for more products
6. We — the taxpayers and districts—foot the bill while companies profit

It's a perfect system if you're selling the solutions, but a disaster if you're teaching or learning.

<div class="callout-tip">
<strong>Let's be honest with each other...</strong> when have you ***EVER*** purchased a 'solution' that actually made your education delivery system less complicated and more effective?
</div>


## Complexity By Design, Not Necessity

I recently asked a fellow Superintendent why experienced teachers need to create detailed lesson plans. The one-word answer? "TEKS."

Let's look at what's happening in our classrooms:

- Our experienced math teachers already know how to teach algebra effectively
- Now they must document exactly how each lesson covers specific standards
- TEKS change constantly, making textbooks and lesson plans quickly outdated
- This creates hours of paperwork that doesn't improve actual teaching
- Software companies offer "alignment tools" that promise compliance
- Our districts buy these expensive programs because we have no choice
- Our teachers then spend valuable time feeding data into these systems instead of teaching

The state creates unnecessary complexity, and the marketplace responds with costly "solutions." In effect, the state has become an unofficial partner in the software and textbook business.


## The Education-Government Complex

This pattern mirrors what President Eisenhower called the "military-Government complex" — where government spending creates industries that then lobby for more spending. In our education system, we see:

```{r ed-complex-table, echo=FALSE, warning=FALSE, message=FALSE}
ed_complex <- data.frame(
  Player = c("Lawmakers", "Education Agencies", "Testing Companies", "Software Vendors", "Consultants", "Professional Development"),
  Role = c(
    "Create mandates requiring complex compliance",
    "Develop detailed regulations and reporting requirements",
    "Create and score tests aligned to state standards",
    "Sell tracking systems for standards and student data",
    "Help districts navigate complex requirements",
    "Train teachers to use the new systems and standards"
  ),
  Profit_Source = c(
    "Political support from companies that benefit",
    "Larger budgets to monitor compliance",
    "Contracts for mandated state testing",
    "Subscriptions to 'compliance' software",
    "High-priced training on navigating mandates",
    "Required training for all teachers"
  )
)

# display the table
knitr::kable(ed_complex, caption = "The Education-Government Complex")

```

Everyone in this system benefits from more complexity - except our teachers and students.


## The Real Costs to Our Educational Mission
This system damages our core educational mission in measurable ways:

1. Teacher burnout: Our best educators leave when paperwork replaces teaching
1. Lost teaching time: Hours spent on compliance mean less time with students
1. Resource drain: Money spent on software could fund smaller classes
1. Focus shift: The goal becomes meeting mandates, not helping students learn
1. Innovation barriers: Creative teaching gets pushed aside by standardization
1. Teacher shortages: Educators entered the profession to teach, not feed computers


## The Empty Promises We Keep Buying

These products seduce districts with appealing promises:

"Save time on lesson planning!"
"Track student progress automatically!"
"Ensure compliance with all state standards!"
"Individualize learning with our algorithm!"

But we've all seen the disappointing reality:

* Teachers still spend hours inputting data into these systems
* The "time-saving" tools create new forms of non-teaching work
* Much of the data collected doesn't actually improve instruction
* The promised individualization rarely delivers meaningful results


## Taking Back Control Together

How can we break this cycle as district leaders? Here are strategies we can implement:

1. Push for simpler standards: Advocate for fewer, clearer standards that experienced teachers can follow without special tools

2. Question every purchase: When vendors pitch solutions, ask "Would this problem exist without the mandate?"

3. Calculate true costs: Add up not just the software price but teacher time feeding these systems

4. Trust our teachers: Give experienced educators more professional judgment in meeting learning goals

5. Explore free alternatives: Many open-source tools can replace expensive proprietary systems


## Questions Every Board Member Should Ask

When faced with the next "essential" education software or textbook purchase, challenge assumptions with these questions:

- "Would we need this if state requirements were simpler?"
- "How much teacher time will this system consume?"
- "Could we meet our goals with existing tools or simpler approaches?"
- "Is this solving an educational problem or just a compliance problem?"
- "What evidence shows this improves actual student learning, not just compliance?"


## Imagine a Different Future

What if we worked together to create:

* Lawmakers and bureaucrats who understand the true costs of their mandates—both financial and human

* Districts that push back against unnecessary complexity by focusing resources where they actually improve teaching and learning

* Communities that demand change at the ballot box

As board members and educational leaders, we must recognize this problem and take action. Our teachers, students, and communities deserve better than a system designed to generate profits rather than learning.




### If you liked this post or found it valuable - share it with others.


### Your Thoughts?

### Share your perspective in the comments below. Use the 'Sign in with GitHub' button to verify you're not a robot.

