---
title: "The Education-Industrial Complex"
subtitle: "How Mandates Create Markets"
description: "(estimated time:   6 mins)"
author: "Dan Swart, CPA (ret)"
date: "April 4, 2025"
output:
  html_document:
    toc: true
    toc_float: true
    theme: paper
    highlight: tango
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```



## The Perfect Money-Making Machine

Here's something important that many people miss: education has become a perfect money-making machine for companies and consultants that sell "solutions" to problems created by government mandates. 

This cycle works like this:

1. Lawmakers create extremely detailed mandates (like the TEKS in Texas)
2. These mandates are too complex for teachers or adminstartors to track manually
3. Companies create expensive "solutions" to help schools keep track of it all
4. Schools must buy these solutions to meet state requirements
5. More mandates create more markets for more products
6. Taxpayers foot the bill while companies profit

It's a perfect system - if you're selling the solutions, not if you're teaching or learning.


## The Problem of Manufactured Complexity

I once asked a Superintendent why very experienced teachers need to continuously create detailed lesson plans.  Came the one-word answer: "TEKS" (these are the state-mandated educational standards for public schools in Texas, outlining what students should know and be able to do in various subjects at each grade level, according to the Texas Education Agency).  

Teaching today carries a load of unnecessary work. Let's break this down:

- An experienced math teacher knows how to teach algebra effectively
- But now they must document exactly how each lesson covers each TEK
- TEKS are constantly changing making text books and lesson plans outdated almost immediately
- This creates hours of paperwork that doesn't improve teaching
- Enter the software and textbook companies with "alignment tools" that promise to save time and guarantee compliance with the TEKS
- Schools buy these expensive programs and textbooks because they have no choice if they hope to stay 'aligned' with the TEKS
- Teachers then spend hours feeding data into these systems, learning how to feed the data, and learning how to access and interpret the reports.

The state creates the complexity and the marketplace responds.  Suddenly, there is money to be made from the 'good intentions' of the education authorities.  For all intents and purposes, the state is in the software and book business.


## The Education-Industrial Complex

This pattern resembles what President Eisenhower called the "military-industrial complex" - where government spending creates industries that then lobby for more spending. In education, we see:

```{r ed-complex-table}
ed_complex <- data.frame(
  Player = c("Lawmakers", "Education Agencies", "Testing Companies", "Software Vendors", "Consultants", "Professional Development"),
  Role = c(
    "Create mandates requiring complex compliance",
    "Develop detailed regulations and reporting requirements",
    "Create and score tests aligned to state standards",
    "Sell tracking systems for standards and student data",
    "Help districts navigate complex requirements",
    "Train teachers to use the new systems and standards"
  ),
  Profit_Source = c(
    "Political support from companies that benefit",
    "Larger budgets to monitor compliance",
    "Contracts for mandated state testing",
    "Subscriptions to 'compliance' software",
    "High-priced training on navigating mandates",
    "Required training for all teachers"
  )
)
```

Each player in this system except teachers and students benefits from more complexity, not less.


## How This Hurts Teaching and Learning

This system has real costs beyond just money:

1. **Teacher burnout**: Experienced teachers quit when paperwork replaces teaching
2. **Lost teaching time**: Hours spent on compliance mean less time with students
3. **Resource drain**: Money spent on software could fund smaller classes or better materials
4. **Focus shift**: The goal becomes meeting mandates, not helping students learn
5. **Innovation barriers**: Creative teaching gets squeezed out by standardization
6. **Teacher shortages**:  Teachers got into the business to teach, not feed computers


## The Seductive Promise of Easy Solutions

The irony is these products are marketed with seductive promises:

- "Save time on lesson planning!"
- "Track student progress automatically!"
- "Ensure compliance with all state standards!"
- "Individualize learning with our algorithm!"

But the reality is different:

- Teachers still spend hours inputting data into the systems
- The "time-saving" tools create new forms of non-classroom work
- Much of the data collected isn't actually useful for teaching
- The promised individualization rarely delivers meaningful results


## Breaking the Cycle

How might we break this destructive cycle? Here are some starting points:

1. **Simplify standards**: Push for fewer, clearer standards that experienced teachers could reasonably follow without special tools

2. **Question new purchases**: When vendors promise solutions, ask "Would this problem exist without the mandate?"

3. **Calculate true costs**: Add up not just the software price but teacher time spent feeding these systems

4. **Trust teachers more**: Give experienced educators more professional judgment in how they meet broad learning goals

5. **Look for free alternatives**: Many open-source and free tools can replace expensive proprietary systems


## Talking Points for Board Members and Administrators

When faced with the next "essential" education software or book purchase, try these questions:

- "Would we need this if the state requirements were simpler?"
- "How much teacher time will this system require to maintain?"
- "Could we meet our goals with existing tools or simpler approaches?"
- "Is this solving a real educational problem or a compliance problem?"
- "What's the evidence this improves actual student learning, not just compliance?"


## Conclusion: Seeing the System Clearly

The first step toward change is seeing the system clearly. The "education-government complex" isn't inevitable - it's a created market that benefits everyone except those it should serve: teachers and students.

By naming this system and understanding how it works, school boards and administrators can start creating advocacy strategies to awaken lawmakers and beauracrats to the astonishing costs for districts and people.  Districts must find ways that push back against unnecessary complexity and focus resources where they actually improve teaching and learning.

As one practical step, boards might require any new purchase to pass a simple test: "Does this primarily solve a teaching problem or a compliance problem?" If it's mainly the latter, perhaps that's where the questioning should begin.
